A thesis draft

### 2018-12-01 res/sections/07-Introduction.tex
Add prof corrections

This new possibility of provisioning a network with specialized, heterogeneous
classes of services, combined with emerging virtualization technologies like
cloud computing, hypervisor-virtualized machine or container-based solutions,
enables the deployment of Virtualized Network Functions (VNFs) within seconds
running on generic purpose or custom hardware. While there may be some
performances degradation, it has been demonstrated that their time-to-deploy and
the possibility of automation they offer are unprecedented~\cite{nguyen2017sdn}.
On top of that, a specialized software entity acts as a MANager and Orchestrator
(MANO) for these VNFs, managing their lifecycle. While most of the current
proposals are based on a hypervisor solution, in this thesis we set the
objective of providing a proof-of-concept implementation of an Infrastructure
Management and Orchestration plane based on the new container-based technology,
called Docker, on a recent orchestrator framework: Kubernetes. Moreover,
building on recent published results of the 3GPP Release 15, we outline a
software architecture capable of end-to-end service orchestration exploiting
container-technology.

@@

This thesis is organized as follows:
\begin{itemize}
 \item Chapter~\ref{chap:background} describes background information about 5G
   and the technologies used to build an ETSI compliant architecture;
 \item Chapter~\ref{chap:rel_wk} gives an insight of correlated works on this
   field;
 \item Chapter~\ref{chap:prjan} explains the analysis performed on the
   requirements provided and the overall project roadmap;
 \item Chapter~\ref{chap:archimpl} provides an in-depth view of the high-level
   architectural implementation;
 \item Chapter~\ref{chap:vnf_ns_impl} gives a brief explanation of the SFC
   implementation;
 \item Chapter~\ref{chap:test} describes the tests done on the platform;
 \item Chapter~\ref{chap:fw} illustrates possible future works;
 \item Chapter~\ref{chap:conclusions} wraps up the overall project experience.
\end{itemize}

>>>>

This new possibility of provisioning a network with specialized, heterogeneous
classes of services, combined with emerging virtualization technologies like
cloud computing, hypervisor-virtualized machine or container-based solutions,
enables the deployment of Virtualized Network Functions (VNFs) within seconds
running on generic purpose or custom hardware. Even though the use
virtualization is sometimes paid in terms of computational overhead, the
technology has demostrated its maturity and the possibilities of automation it
offers are unprecedented~\cite{nguyen2017sdn}. On top of that, a specialized
software entity acts as a MANager and Orchestrator (MANO) for these VNFs,
managing their lifecycle. While most of the current proposals are based on a
hypervisor solution, in this thesis we set the objective of providing a
proof-of-concept implementation of an Infrastructure Management and
Orchestration plane based on the new container-based technology, called Docker,
on a recent orchestrator framework: Kubernetes. Moreover, building on recent
published results of the 3GPP Release 15, we outline a software architecture
capable of end-to-end service orchestration exploiting container-technology.

@@

This thesis is organized as follows:
\begin{itemize}
 \item Chapter~\ref{chap:background} provides some background information about
   5G and the technologies used to build an ETSI compliant architecture;
 \item Chapter~\ref{chap:rel_wk} discusses some prior work of relevance in the
   context;
 \item Chapter~\ref{chap:prjan} provides a detailed analysis of the domain,
   fixing some design requirments, stating the overall project roadmap.
 \item Chapter~\ref{chap:archimpl} provides an in-depth view of the high-level
   architectural implementation;
 \item Chapter~\ref{chap:vnf_ns_impl} gives a brief explanation of the SFC
   implementation;
 \item Chapter~\ref{chap:test} discusses the experimental setting along with
   some obtained preliminary results
 \item Chapter~\ref{chap:fw} illustrates possible future works;
 \item Chapter~\ref{chap:conclusions} wraps up the overall project experience.
\end{itemize}

### 2018-12-01 res/sections/08-Background.tex
Change paper images
Add prof corrections

In this chapter we are going to explain the most important concepts of the
upcoming 5G standard, talking about the Virtual Network Functions (VNFs), the
Network Slice (NS) concept and how first experiments with these new innovations
already begun. Finally we are going to briefly introduce and explain the
technologies involved and their role.

@@

\subsubsection{Network slicing}
\label{chap:background:sec:5g:sub:ns}
The role of a \emph{network slice} in a 5G architecture is to specifically
handle the Control-plane of a particular service (e.g. smartphones traffic,
autonomous driving, massive IoT), deploying resources in a manner that assure
the required latency, security and reliability. While some very peculiar and
legacy service could require specific hardware, the common resources between
services could be shared in a virtualized way, providing auto-scaling
capabilities in services that are under heavy network pressure.

@@

This thesis started as a part of the VIBES project~\cite{vibesesa}, whose main
requirement is to enhance the performance of end-to-end IP based that involves
satellite connections. To reach this goal, the project specifications suggest to
exploit the 5G incoming technology and use the NFV-MANO architecture to perform
first packets elaboration and performance improvement and finally TCP/IP
satellite chunk optimization with the Performance Enhancing Proxy (PEP). The
VIBES project proposed five technical requirements:
\begin{enumerate}
  \item Analysis of the applicability of current and new Internet protocols in
  the proposed VNF-PEP architecture;
  \item Implementation of a VNF-PEP prototype;
  \item Building of a PoC test platform;
  \item VNF-PEP validation and performance tests on 5G use-cases;
  \item Demonstration test-bed management.
\end{enumerate}

@@

\section{Virtualization}
It is remarkable how cloud computing has surged in the recent years. This is due
the numerous features it carries, such as a great way to isolate processes and
mitigate possible attacks from malicious users. On top of that, it allows to run
virtual machines with different configurations, without bringing any changes to
the host machine. Indeed, the virtualization concepts refer to the property of a
machine to be virtualized. To achieve this, the components that forms a common
computer need to be virtualized too. Thus, these assets are recreated in this
virtualized manner, e.g. storage devices and computer network
resources~\cite{liu2014research}. At the time of writing, there are two common
virtualization options: the virtualization through Hypervisor (defined also as
hypervisor-based) and the container through virtualization.

@@

\begin{figure}[t]
  \centering
  \includegraphics[scale=2]{etsi_arch}
  \caption[Network Function Virtualization reference architecture]{Network
    Function Virtualization reference architecture~\cite{etsi2013gs}}
  \label{chap:background:img:etsi_arch}
\end{figure}

>>>>

This chapter provides some background information related to 5G technology
confined to the domain under scrutiny. We start by introducing the concept the
ETSI NFV- Management and Orchestration (MANO) proposal, outlining outlining key
functional components and their modus operandi. Next, we provide some background
information on the Software Defined Networking (SDN) paradigm and its
data/control plane softwarization effort. Concluding, we introduce the Service
Function Chaining paradigm, arguing how all the above mentioned technology could
be employed in synergy in order to implement an end-to-end service delivery
scheme.

@@

\subsubsection{Network slicing}
\label{chap:background:sec:5g:sub:ns}

Network Slicing is a 5G related concept referring to the possibility of
provisioning an end-to-end optimized service within the network, that responds
to particular Quality of Service requirements (e.g. low latency, high
throughput). The service under consideration might be constituted by one or more
functional, either virtualized of physical components, all controlled and
management by a logically centralized network brain. The service provider
exposes a set of well-defined APIs to the consumer, through which it is possible
to define and instantiate a fully fledged service whose deployment is enacted on
a shared infrastructure layer. It is noteworthy to point out that no exact
definition of a network slice exists, and different approaches on how to achieve
it have been advanced.

While some very peculiar and legacy service could require specific hardware, the
common resources between services could be shared in a virtualized way,
providing auto-scaling capabilities in services that are under heavy network
pressure.

@@

VIBeS is a concrete example of a 5G project combining the use of SDN and NFV
technology, working in concert to provide specialized network slices for several
envisioned use-cases~\cite{vibesesa}. VIbES goals are ambitious, aiming to
demonstrate the practical convergence of satellite and 5G in a unified frame. To
reach this goal, the project specifications suggest to exploit the 5G incoming
technology and use the NFV-MANO architecture to perform first packets
elaboration and performance improvement and finally TCP/IP satellite chunk
optimization with the Performance Enhancing Proxy (PEP). The VIBES project
proposed five technical requirements:
\begin{enumerate}
  \item Analysis of the applicability of current and new Internet protocols in
  the proposed VNF-PEP architecture;
  \item Implementation of a VNF-PEP prototype;
  \item Building of a PoC test platform;
  \item VNF-PEP validation and performance tests on 5G use-cases;
  \item Demonstration test-bed management.
\end{enumerate}

@@

\section{Virtualization Technology}
It is remarkable how cloud computing has surged in the recent years. This is due
the numerous features it carries, such as a great way to isolate processes and
mitigate possible attacks from malicious users. On top of that, it allows to run
virtual machines with different configurations, without bringing any changes to
the host machine. Indeed, the virtualization concepts refer to the property of a
machine to be virtualized. To achieve this, the components that forms a common
computer need to be virtualized too. Thus, these assets are recreated in this
virtualized manner, e.g. storage devices and computer network
resources~\cite{liu2014research}. At the time of writing, there are two common
virtualization options: the virtualization through Hypervisor (defined also as
hypervisor-based) and the container through virtualization.

@@

\begin{figure}[t]
  \centering
  \includegraphics[scale=0.5]{etsi_arch}
  \caption[Network Function Virtualization reference architecture]{Network
    Function Virtualization reference architecture~\cite{etsi2013gs}}
  \label{chap:background:img:etsi_arch}
\end{figure}

### 2018-12-01 res/sections/09-RelatedWork.tex
Add prof corrections

The possibility to solve the increasing network capacity necessities and to make
the network infrastructure lighter and faster to deploy has rouse the interests
of many internet service providers (ISP).

@@

In literacy, several NFV implementations and Network Slicing solutions have been
proposed and some of them will be discussed in the following chapter. For
example, an opportunity identified in~\cite{nogales2018nfv} shows that ETSI MANO
implementations can be deployed inside Unmanned Aerial Vehicles (UAVs), giving
the possibility to rapidly deploy network connectivity in areas where the common
communication systems are temporarily unavailable. This can be very useful in
places where natural disasters occurred and there is the necessity to restore
the communications within a very short span of time. On top of that, this system
is flexible enough to be able to provide different kind of services, in addition
to simple broadband connectivity. The only downside of this approach are the
constrained system sources: the authors had to install the VIM (in particular,
Openstack) in a separated machine, located on the ground. A solution based only
on containerization, instead, could lead to interesting results, since
Kubernetes does not require too much resources \textit{per se} and in literature
is possible to find solutions of Kubernetes installations performed on machines
with the same hardware installed on the UAVs~\cite{pahl2016container}.

>>>>

As argued, service differentiation and network management are the swiss army
kife of network operators. Through the rest of this chapter we discuss some
prior work in the context of ETSI-NFV, providing end-to-end management
capabilities into the network.

@@

In literature, several NFV implementations and Network Slicing solutions have been
proposed and some of them will be discussed in the following chapter. For
example, an opportunity identified in~\cite{nogales2018nfv} shows that ETSI MANO
implementations can be deployed inside Unmanned Aerial Vehicles (UAVs), giving
the possibility to rapidly deploy network connectivity in areas where the common
communication systems are temporarily unavailable. This can be very useful in
places where natural disasters occurred and there is the necessity to restore
the communications within a very short span of time. On top of that, this system
is flexible enough to be able to provide different kind of services, in addition
to simple broadband connectivity. The only downside of this approach are the
constrained system sources: the authors had to install the VIM (in particular,
Openstack) in a separated machine, located on the ground. A solution based only
on containerization, instead, could lead to interesting results, since
Kubernetes does not require too much resources \textit{per se} and in literature
is possible to find solutions of Kubernetes installations performed on machines
with the same hardware installed on the UAVs~\cite{pahl2016container}.

### 2018-12-01 res/sections/10-ProjectAnalysis.tex
Add prof corrections

As described in chapter~\ref{chap:intro}, the creation of a full ETSI MANO
compliant architecture, following the suggestions described in RFC 7665,
requires the integration of multiple tools (some of them have been already
briefly introduced in~\ref{chap:intro}). The tools were chosen after their
analysis, where pro and cons have been studied and compared to the necessity of
this thesis. Since the project has been developed with the cloud provided by the
University of Padua, Openstack (which a product analysis can be found later
in~\ref{chap:prjan:sec:openstack}) has been used as base for our deployments.

\section{Available technologies in the market}
\label{chap:prjan:sec:tech}
Before starting any effective work on the project it has been decided to 
perform an analysis of the possible technologies available in the market, 
performing a choice between them and exploiting their possible ``weak 
points''. Proceeding with a top-down approach, the frameworks were chosen from 
the one that had the biggest role in the test-bed to conclude with the 
less-important components.
\subsection{VIM}
A critical component, as the MANO one, is the VIM. Without it, network resources
can not be managed, and since our project was focused on dynamically allocating
containerized applications on cloud platforms, we decided to start analyzing the
right framework for this component first.

>>>>

As discussed through Chapter ~\ref{chap:intro}, the implementation of an
ETSI-compliant MANO architecture, contemplating service chain proovisioning
capabilities as mandated by RFC 7665, requires the integration of multiple
tools. The tools were chosen after their analysis, where pro and cons have been
studied and compared to the necessity of this thesis. Since the project has been
developed within the cloud environment provided by the Department of
Mathematics, University of Padua, we initially relied on Openstack (which a
product analysis can be found later in~\ref{chap:prjan:sec:openstack}) as the
default infrastructure manager choice.

\section{Available technologies in the market}
\label{chap:prjan:sec:tech}
Before starting any effective work on the project it has been decided to 
perform an analysis of the possible technologies available in the market, 
performing a choice between them and exploiting their possible ``weak 
points''. Proceeding with a top-down approach, the frameworks were chosen from 
the one that had the biggest role in the test-bed to conclude with the 
less-important components.

\subsection{Virtualized Infrastructure Manager}
A critical component, as the MANO one, is the VIM. Without it, network resources
can not be managed, and since our project was focused on dynamically allocating
containerized applications on cloud platforms, we decided to start analyzing the
right framework for this component first.

### 2018-12-01 res/sections/15-ArchitecturalImplementation.tex
Add prof corrections

In this chapter we will dig into the actual architectural proposal, seeing all
the problems related to it and how they were solved. We will show all the
various attempts made, and we will discussion about the technological issues and
changes we had to perform.

\section{Architecture introduction}
Before describing the real implementation, we are going to briefly introduce
some of its functional components and the way they work.

\subsection{A two-level orchestration framework}
\begin{figure}[t]
  \centering
  \includegraphics[scale=0.6]{Architecture}
  \caption[Architecture schema depicting the expected
    implementation]{Architecture schema depicting the expected implementation.
    It is possible to see how Openbaton interacts with the various components in
    the architecture, and how RabbitMQ is used as a message broker between the
    different parts to enable message exchanging.}
  \label{chap:archimpl:sec:fistattempt:img:architecture}
\end{figure}

@@

The lack of these two management requires a service aware software, that has the
task to perform service lifecycle management and monitoring via the API offered
by the ETSI MANO. In this context, is possible to identify two different types
of orchestrators, as represented in
Figure~\ref{chap:archimpl:sec:fistattempt:img:architecture}:
\begin{itemize}
\item a high layer, service-aware;
\item a low layer, service-unaware.
\end{itemize}
Both the layers would use Openbaton. The messages would be exchanged through
RabbitMQ, using the Advance Message Queuing Protocol (AMQP).

@@

\section{Second attempt - revision one}

@@

\subsection{Harbor}
\label{chap:archimpl:sec:secondattempt:sub:harbor}
Harbor grew considerably during the project development, as it revealed to be a
key component of our architecture. Harbor does not have a web interface, but as
already written it has a RESTful API that allows users to interact with it using
CRUD operations. It offers functionalities similar to Openbaton: possibility to
save and retrieve VNFs and SFCs definitions. When an SFC is saved, a consistency
check is performed to assure that all the VNFs have already been defined in
Harbor. On an SFC deployment, with a mechanism similar to the Openbaton one, all
the associated VNFs are set up too, making the whole operation of launching a
SFC atomic from the user point-of-view. Since this architecture component grew
until it became a Kubernetes MANO, we added a feature that, in our opinion,
Openbaton lacks: smart resource utilization. Harbor, in fact, during an SFC
deletion, checks if any related VNF remains orphan (which means without any
SFC using it). In that case the VNF will enter in a state of pruning: after five
minutes, if the VNF has not be re-utilized by any other SFC it will be deleted.
This mechanism assure faster deployments in environments where NSs are
continuously taken up and down.

@@

\subsection{Final architecture structure}
At the end, we were able to create a scalable architecture that followed the
ETSI MANO standard, with Kubernetes as VIM and Harbor has MANO. We faced
additional challenges regarding the system scalability, that we solved using the
Roulette controller. We eased the creation of VNFs with Astaire, and we created
proper ingress and egress points with Ironhide, that supported scalability
issues and performed a protocol-based packet classification. Is possible to see
a final architectural schema in
Figure~\ref{chap:archimpl:sec:secondattempt:img:attempt2v1total}.

>>>>

In this chapter we dig into the actual architectural proposal, seeing all the
problems related to it and how they were solved. We will show all the various
attempts made, and we will discussion about the technological issues and changes
we had to perform.

\section{Architecture introduction}

Before devling into the details of the actual implementation, we briefly
introduce some of its functional components and the way they work.

\subsection{A two-level orchestration framework}
\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{Architecture}
  \caption[Architecture schema depicting the expected
    implementation]{Architecture schema depicting the expected implementation.
    It is possible to see how Openbaton interacts with the various components in
    the architecture, and how RabbitMQ is used as a message broker between the
    different parts to enable message exchanging.}
  \label{chap:archimpl:sec:fistattempt:img:architecture}
\end{figure}

@@

The lack of these two management requires a service aware software, that has the
task to perform service lifecycle management and monitoring via the API offered
by the ETSI MANO. In this context, is possible to identify two different types
of orchestrators, as represented in
Figure~\ref{chap:archimpl:sec:fistattempt:img:architecture}:
\begin{itemize}
\item a high layer, service-aware;
\item a low layer, service-unaware.
\end{itemize}
Both layers rely on the Openbaton NFVO. The messages would be exchanged through
RabbitMQ, using the Advance Message Queuing Protocol (AMQP).

@@

\section{Final Design: End-to-end MANO}

@@

\subsection{Harbor: extending the capabilities}
\label{chap:archimpl:sec:secondattempt:sub:harbor}
Harbor grew considerably during the project development, as it revealed to be a
key component of our architecture. Harbor does not have a web interface, but as
already written it has a RESTful API that allows users to interact with it using
CRUD operations. It offers functionalities similar to Openbaton: possibility to
save and retrieve VNFs and SFCs definitions. When an SFC is saved, a consistency
check is performed to assure that all the VNFs have already been defined in
Harbor. On an SFC deployment, with a mechanism similar to the Openbaton one, all
the associated VNFs are set up too, making the whole operation of launching a
SFC atomic from the user point-of-view. Since this architecture component grew
until it became a Kubernetes MANO, we added a feature that, in our opinion,
Openbaton lacks: smart resource utilization. Harbor, in fact, during an SFC
deletion, checks if any related VNF remains orphan (which means without any SFC
using it). In that case the VNF will enter in a state of pruning: after five
minutes, if the VNF has not be re-utilized by any other SFC it will be deleted.
This mechanism assure faster deployments in environments where NSs are
continuously taken up and down.

@@

\subsection{Discussion}
At the end, we were able to create a scalable architecture that followed the
ETSI MANO standard, with Kubernetes as VIM and Harbor has MANO. We faced
additional challenges regarding the system scalability, that we solved using the
Roulette controller. We eased the creation of VNFs with Astaire, and we created
proper ingress and egress points with Ironhide, that supported scalability
issues and performed a protocol-based packet classification. Is possible to see
a final architectural schema in
Figure~\ref{chap:archimpl:sec:secondattempt:img:attempt2v1total}.
